---
title: "Last Linear Regression Discussion"
output: html_notebook
---



# Goals

- Continue exploring linear regression.
- Answer questions.


## Task 0:

Next week we will turn our attention to odds, odds ratios, and logistic
regression. Therefore, I want to take few minutes and answer any questions you
all may have about linear regression, data transformations, etc.

This will be a group discussion.



# Setup

I am creating models with very similar names, like `lm_simple`. PLEAES clear
your global environment before working on this lab so you don't accidentally use
a model created during the lecture in this lab by accident.

If you aren't sure how to do this, please ask.

```{r}
#| label: setup
#| message: false
#| warning:false

library(Stat2Data)
library(janitor)
library(hexbin)
library(lubridate)
library(tidymodels)
library(tidyverse)

options(scipen = 999)

diamonds <- read_csv("data/diamonds.csv")

```



# Data - Diamonds

## Data

### Description:

A dataset containing the prices and other attributes of almost 54,000 diamonds.

### Format:

A data frame with 53940 rows and 10 variables:

- price: price in US dollars ($326-$18,823)
- carat: weight of the diamond (0.2-5.01)
- cut: quality of the cut (Fair, Good, Very Good, Premium, Ideal)
- color: diamond colour, from D (best) to J (worst)
- clarity: a measurement of how clear the diamond is (I1 (worst),
     SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))
- x: length in mm (0-10.74)
- y: width in mm (0-58.9)
- z: depth in mm (0-31.8)



# Lab Tasks

- Remember, a model without a plot is a dangerous thing.


## Task 01:

Before attempting this task, please review the data dictionary/information
above.

- **Question:** Is there a linear relationship between the weight of a diamond
  and the price of the diamond?
  - HINT: This involves ggplot.
- **Answer:** 
  - I am looking for a yes/no here.


YOUR CODE HERE!


## Task 02:

OK, I'll go ahead and admit it, the answer to Task 01 is NO. If you answered
yes, I want you to go back and look at your plot very carefully before
proceeding and maybe we should discuss.

Fortunately, we can fix this. Price changes rapidly as a function of weight. In
fact, it changes exponentially. This happens often in things like price. But, if
you look, we also quite a bit of vertical banding in the data.

```{r}

diamonds |>
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = 0.25)

diamonds |>
  ggplot(aes(x = carat, y = price)) +
  geom_hex()

```

- **Question:** Do you see any banding in the data which can be explained by our
  categorical variables?
- **Answer:** 
    - Answer should be cut, color, clarity and a plot(s) to support your answer.
- Helpful Hints:
    - My code sample above shows you how to change the alpha level of a
      `geom_point()`, which will help you not get overwhelmed by the large
      number of data points in diamonds.
    - I also introduced a new function, `geom_hex()` which will help you.
        - Use categorical features in your `aes()` with something like: 
        `fill = foo`.
    - Explore this data using both methods.
  

YOUR CODE HERE!


## Task 03:

But, before we we actually develop a linear regression, we have to do SOMETHING
about that price vs carat curve.

```{r}

diamonds |>
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = lm) +
  ylim(0, 20000)

```

Look at the linear regression. The distribution of our residuals is not evenly
distributed. We have more residuals to the upper left of our line than we do to
the upper right of our regression line, because carat and cost do not have a
simple linear relationship.

Fortunately, we can fix this using the log function. Create a new column,
`price_log` using the following code and then graph the relationship again.

```{r}

diamonds <-
  diamonds |>
  mutate(
    price_log = log(price),
    carat_log = log(carat)
  )

diamonds |>
  ggplot(aes(x = carat_log, y = price_log)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = lm) +
  ylim(4.8, 10.2)

```

I want to take a moment to review what we have done.

```{r}

diamonds |> head(10) |> select(price, price_log, carat, carat_log)

diamonds |> summarize(min(price), min(price_log), max(price), max(price_log))

diamonds |> summarize(min(carat), min(carat_log), max(carat), max(carat_log))

```


- **Question:** While not perfect, taking the log of both carat and price gives
  us a much more linear relationship. In the code sample above, I take the log
  of both carat and price. Was this necessary (yes/no) and why (use ggplots to
  support your answer).
- **Answer:** 
- Helpful Hint:
    - The original price and carat data is still available in `diamonds`.
    - We are using the natural log function, which you can read about here:
    https://en.wikipedia.org/wiki/Natural_logarithm


YOUR CODE HERE!


## Task 04:

- **Question:** Create a model, `lm_simple` which models `price_log` as a
  function of `carat_log`.
- **Answer:** What is the coefficient for carat_log?
- Reasons to use the natural log in linear regression:
    - Linearize Data: Many real-world relationships between variables are
      multiplicative rather than additive. Taking the natural log of variables
      can transform these multiplicative relationships into additive ones,
      making them more suitable for linear regression.
    - Addressing Skewness: Natural logs can also help make skewed data more
      normally distributed, which is a desirable characteristic for linear
      regression.
    - Handling Heteroscedasticity: By transforming variables, natural logs can
      sometimes help reduce heteroscedasticity (unequal variance of errors) in
      the residuals, which can improve the reliability of the regression model.
    - Interpreting Coefficients: When using natural logs, the interpretation of
      regression coefficients changes. A coefficient on a natural-log scale
      represents the approximate percentage change in the dependent variable for
      a one-unit change in the independent variable.


YOUR CODE HERE!


Remember, our coefficients are no longer coefficients of actual carats or
prices.


## Task 05:

The opposit of the `log()` function is the exponent function, `exp()`. Can you
use augment and your dplyr skills to transform your predicted price_log into a
predicted price output and plot the density plot of your predicted prices?

- **Question:** Draw a density plot of the residuals from lm_simple.
- **Answer:** 
- Helpful Hints:
    - In the lectures, I change the name of .pred and .resid.
    - You don't need to do this. I know, it can get confusing.
    - I recommend thinking this task through carefully, creating the new column
      and plotting it directly without creating an entirely new data set.


YOUR CODE HERE!


## Task 06:

Please signal to me when your group gets to Task 06.
