---
title: "Ratios"
output: html_notebook
---


# Setup

```{r message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
strokes <- read_csv("data/strokes.csv") %>% filter(gender != "Other")
```



# Data - Strokes

Source: [Kaggle Stroke Prediction Dataset](https://www.kaggle.com/fedesoriano/stroke-prediction-dataset)

> According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths. This dataset is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relavant information about the patient.

```{r paged.print=TRUE}
strokes
```

1. id: unique patient identifier
2. gender: "Male", "Female" or "Other"
3. age: age of the patient
4. hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
5. heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
6. ever_married: "No" or "Yes"
7. work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
8. residence_type: "Rural" or "Urban"
9. avg_glucose_level: average glucose level in blood
10. bmi: body mass index
11. smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12. stroke: 1 if the patient had a stroke or 0 if not
*Note: "Unknown" in smoking_status means that the information is unavailable for this patient



# Lab

## Task 01

- Question: The data includes some information about the employment of the patients. Which work type is the most prevalent in this data set?
- Answer:

To prevent confusion, your options below are identical to what you should have found in your data. In Canvas complete this task with one of the following:

    1. children
    2. Govt_job
    3. Never_worked
    4. Private
    5. Self-employed

Helpful Hints: The entries in the `work_type` column are formatted oddly, but that's OK. There's always _something_ funny about a data set.

```{r}
## YOUR CODE HERE!!!

```

## Task 02

- Question: Use R to calculate the percent of patients who are children (based on work type/Task 01).
- Answer: 

Helpful Hints:

- In other words, divide the work_type count by the total.
- And yes, you CAN do this without R, but the point is to do this using R.
- And the easiest way is to calculate the percent for all of the work types.

```{r}
## YOUR CODE HERE!!!

```


## Task 03

- Question: Which patients are more likely to have had a stroke, men or women? 
- Answer:

Helpful Hints:

- We want to calculate the percentage of patients with a stroke.
    - This is just like what we did with the titanic data, but we calculated the percent of passengers who died.
    - But it is the EXACT same idea.

```{r}
## YOUR CODE HERE!

```


## Task 04

- Question: Do men with hypertension have a higher risk of stroke than women with hypertension?
- Answer: 

- This is ALMOST the same question.
- All we need to do is to filter down to those with hypertension.
- Filter should come before group by.

```{r}
## YOUR CODE HERE!!!

```


## Task 05

- Question: How many patients are over the age of 65?
- Answer:

```{r}
## YOUR CODE HERE!

```


## Task 06

- Question: What is the risk ratio for stroke for men compared to women?
- Answer: 

Helpful Hint:

- It may help to calculate this all as two steps.
- First calculate the risk of stroke for men and women.
    - Save this in a table called risk.
- Then calculate the risk ratio.
- And remember, if you can't figure out the fancy code for the risk ratio but you can figure out the gender-specific risk, you can just use R like a calculator and calculate the risk ratio manually.

```{r}
## YOUR CODE HERE!!!

```

## Task 07

- Question: What is the odds ratio for for stroke for men compared to women?
- Answer: 


```{r}
## YOUR CODE HERE!

```

PLEASE PLEASE PLEASE notice that the odds ratio and risk ratio are nearly identical with this data. This is because strokes are rare while drowning on the Titanic was not.

## Task 08

- Question: Demonstrate that the same odds ratio can be obtained via logistic regression. Build a model for risk stroke as a function of gender. What is the coefficient for genderMale?
- Answer:

```{r}
## YOUR CODE HERE!!!

```

For more on this topic: [There's Nothing Odd About Th e Odds Ratio](https://www.statisticssolutions.com/theres-nothing-odd-about-the-odds-ratio-interpreting-binary-logistic-regression/)

Logistic regression allows us to to model more complex models and relationships than we would ever want to model by hand.

