---
title: "Last Linear Regression Discussion"
---

```{r}
#| label: setup
#| message: false
#| warnings: false

library(tidymodels)
library(tidyverse)

options(scipen = 999)

penguins <- read_csv("data/penguins.csv")

```



Linear Regression - Not always simple
================================================================================

- For this discussion, we will return to the penguins data set.
- Let's predict flipper length as a function of body mass.


```{r}

# Observe that I am removing any row where body_mass_g or flipper_length_mm
# is NA.
penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = lm)

```

- This looks like a simple linear regression.
- But, how many rows did I remove?

```{r}

penguins |>
  summarize(
    n_rows = n(),
    missing_rows = sum(case_when(
      is.na(flipper_length_mm)~1,
      is.na(body_mass_g)~1,
      TRUE~0)),
    p_missing = 100*missing_rows/n_rows
    )

```

- Since all I'm doing is building a linear regression, there isn't necessarily a
  simple way to calculate how many rows is too many to drop.
      - That said, my rule of thumb is that if I am dropping more than a few
        percent, I need to think about what I'm doing.
- Next, let's build the linear regression model.
- I am also going to plot the residuals.


```{r}

# Build the model!
lm_simple <-
  linear_reg() |>
  fit(
    flipper_length_mm~body_mass_g,
    data = penguins |> filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g))
  )

# Store residuals in our data set.
penguins <-
  lm_simple |>
  augment(penguins) |>
  select(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    sex,
    year,
    body_mass_g,
    pred_simple = .pred,
    resid_simple = .resid
  )

# Model results
output <-
  penguins |>
  filter(!is.na(resid_simple)) |>
  summarize(
    avg_resid = mean(resid_simple),
    sd_resid = sd(resid_simple),
    lower_95 = avg_resid - 2*sd_resid,
    upper_95 = avg_resid + 2*sd_resid
  )

# Plots my lower and upper 95%
penguins |>
  filter(!is.na(resid_simple)) |>
  # Remember to use + to pipe in ggplot!
  ggplot(aes(x = resid_simple)) +
  geom_density() +
  geom_vline(aes(xintercept = output$lower_95), color = "red") +
  geom_vline(aes(xintercept = output$upper_95), color = "red")

```

- This is working fairly well although our residuals are a little skewed.
- But I wonder how we could tighten this model up.
- I wonder how species affects the relationship between body weight and flipper.


```{r}

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point()

```

- Although we could not see it earlier, we do have three distinct species and
  the Gentoos (blue) are quite separate from the Adelies and Chinstraps.
- So let's add species to our model.
- Up to this point, I've pointed out that ggplot would simply show us the model
this way:


```{r}

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  geom_smooth(method = lm)

```

There are TWO things I want you to observe in this plot.

1. Each species has a slightly different slope. 
    - The slope for Adelie and Gentoo appears similar.
    - The slope for Chinstrap is clearly steeper.
2. Each species wilkl have a different intercept value with Gentoo having the
   highest intercept value.
   
However, R is a flexible tool and we can model this additional feature,
`species`, in several different ways. We will use the formula operators to create the following models:

- `+` to create `lm_intercept`
- `:` to create `lm_slope`
- `lm_combined` will combine the best apects of both!

There are a few other ways to model this, but this is enough for today's
discussion. But first I want to draw a scatter plot and super-impose our model
over our data. In doing so, we will disable ggplot's default plot.

- In the graphs that follow, I will overlay the new model on top of the
  regression line from `lm_simple`.

```{r}

tidy(lm_simple)

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  geom_abline(
    aes(
      intercept = lm_simple$fit$coefficients[1],
      slope = lm_simple$fit$coefficients[2],
      ),
    color ="darkblue")

```

Let's build a model using `+` in our formula, which we will call
`lm_intercept`. In this model, each species will have:

- a different intercept
- the same slope


```{r}

# Overall, this is very similar to what we did before.

# A new model.
lm_intercept <-
  linear_reg() |>
  fit(
    # This is the most important line in this code chunk!
    flipper_length_mm~body_mass_g + species,
    data = penguins |> filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species))
  )

penguins <-
  # Pull from the new model.
  lm_intercept |>
  augment(penguins) |>
  select(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    sex,
    year,
    body_mass_g,
    # Keeping the old model info.
    pred_simple,
    resid_simple,
    pred_intercept = .pred,
    resid_intercept = .resid
  )

output <-
  penguins |>
  filter(!is.na(resid_intercept)) |>
  summarize(
    avg_resid = mean(resid_intercept),
    sd_resid = sd(resid_intercept),
    lower_95 = avg_resid - 2*sd_resid,
    upper_95 = avg_resid + 2*sd_resid
  )

penguins |>
  filter(!is.na(resid_intercept)) |>
  # Remember to use + to pipe in ggplot!
  ggplot(aes(x = resid_intercept)) +
  geom_density() +
  geom_vline(aes(xintercept = output$lower_95), color = "red") +
  geom_vline(aes(xintercept = output$upper_95), color = "red")

```

- Our model is a bit more centered, but what have we done?
- Hopefully, we have given each species a different intercept value.


```{r}

colors <- scales::hue_pal()(3)

tidy(lm_intercept)

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  # Our simple model regression line.
  geom_abline(
    aes(
      intercept = lm_simple$fit$coefficients[1],
      slope = lm_simple$fit$coefficients[2],
      ),
    color ="darkblue") +
  # Regression line for Adelie penguins
  geom_abline(
    aes(
      intercept = lm_intercept$fit$coefficients[1],
      slope = lm_intercept$fit$coefficients[2],
      ),
    color = colors[1],
    linewidth = 1.25) +
  # Regression line for Chinstrap penguins
  geom_abline(
    aes(
      intercept = lm_intercept$fit$coefficients[1] + lm_intercept$fit$coefficients[3],
      slope = lm_intercept$fit$coefficients[2],
      ),
    color = colors[2],
    linewidth = 1.25) +
  # Regression line for Gentoo penguins
  geom_abline(
    aes(
      intercept = lm_intercept$fit$coefficients[1] + lm_intercept$fit$coefficients[4],
      slope = lm_intercept$fit$coefficients[2],
      ),
    color = colors[3],
    linewidth = 1.25)

```

Now, let's give each species a different slope. To do so let's build a model
using `:` in our formula, which we will call `lm_slope`. In this model, each
species will have:

- a different slope
- the same intercept


```{r}

# Overall, this is very similar to what we did before.

# A new model.
lm_slope <-
  linear_reg() |>
  fit(
    # This is the most important line in this code chunk!
    flipper_length_mm~body_mass_g:species,
    data = penguins |> filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species))
  )

penguins <-
  # Pull from the new model.
  lm_slope |>
  augment(penguins) |>
  select(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    sex,
    year,
    body_mass_g,
    # Keeping the old model info.
    pred_simple,
    resid_simple,
    pred_intercept,
    resid_intercept,
    pred_slope = .pred,
    resid_slope = .resid
  )

output <-
  penguins |>
  filter(!is.na(resid_slope)) |>
  summarize(
    avg_resid = mean(resid_slope),
    sd_resid = sd(resid_slope),
    lower_95 = avg_resid - 2*sd_resid,
    upper_95 = avg_resid + 2*sd_resid
  )

penguins |>
  filter(!is.na(resid_slope)) |>
  # Remember to use + to pipe in ggplot!
  ggplot(aes(x = resid_intercept)) +
  geom_density() +
  geom_vline(aes(xintercept = output$lower_95), color = "red") +
  geom_vline(aes(xintercept = output$upper_95), color = "red")

```

- Our model is a bit more centered, but what have we done?
- Hopefully, we have given each species a different slope.


```{r}

tidy(lm_slope)

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  # Our simple model regression line.
  geom_abline(
    aes(
      intercept = lm_simple$fit$coefficients[1],
      slope = lm_simple$fit$coefficients[2],
      ),
    color ="darkblue") +
  # Regression line for Adelie penguins
  geom_abline(
    aes(
      intercept = lm_slope$fit$coefficients[1],
      slope = lm_slope$fit$coefficients[2],
      ),
    color = colors[1],
    linewidth = 1.25) +
  # Regression line for Chinstrap penguins
  geom_abline(
    aes(
      intercept = lm_slope$fit$coefficients[1],
      slope = lm_slope$fit$coefficients[3],
      ),
    color = colors[2],
    linewidth = 1.25) +
  # Regression line for Gentoo penguins
  geom_abline(
    aes(
      intercept = lm_slope$fit$coefficients[1],
      slope = lm_slope$fit$coefficients[4],
      ),
    color = colors[3],
    linewidth = 1.25)

```

- Although not dramatically different, each species has a different slope.

Finally, we will build a model using `+` and `:` in our formula, which we will
call `lm_combined`. In this model, each species will have:

- a different intercept
- a different slope
- reproducing what ggplot gave us


```{r}

# A new model.
lm_combined <-
  linear_reg() |>
  fit(
    # This is the most important line in this code chunk!
    flipper_length_mm~body_mass_g:species + species,
    data = penguins |> filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species))
  )

penguins <-
  # Pull from the new model.
  lm_combined |>
  augment(penguins) |>
  select(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    sex,
    year,
    body_mass_g,
    # Keeping the old model info.
    pred_simple,
    resid_simple,
    pred_intercept,
    resid_intercept,
    pred_combined = .pred,
    resid_combined = .resid
  )

output <-
  penguins |>
  filter(!is.na(resid_combined)) |>
  summarize(
    avg_resid = mean(resid_combined),
    sd_resid = sd(resid_combined),
    lower_95 = avg_resid - 2*sd_resid,
    upper_95 = avg_resid + 2*sd_resid
  )

penguins |>
  filter(!is.na(resid_combined)) |>
  # Remember to use + to pipe in ggplot!
  ggplot(aes(x = resid_intercept)) +
  geom_density() +
  geom_vline(aes(xintercept = output$lower_95), color = "red") +
  geom_vline(aes(xintercept = output$upper_95), color = "red")

```

- Our model is a bit more centered, but what have we done?
- Observe that we aren't really shrinking the standard deviation any here.


```{r}

tidy(lm_combined)

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  # Our simple model regression line.
  geom_abline(
    aes(
      intercept = lm_simple$fit$coefficients[1],
      slope = lm_simple$fit$coefficients[2],
      ),
    color ="darkblue") +
  # Regression line for Adelie penguins
  geom_abline(
    aes(
      intercept = lm_combined$fit$coefficients[1],
      slope = lm_combined$fit$coefficients[4],
      ),
    color = colors[1],
    linewidth = 1.25) +
  # Regression line for Chinstrap penguins
  geom_abline(
    aes(
      intercept = lm_combined$fit$coefficients[1] + lm_combined$fit$coefficients[2],
      slope = lm_combined$fit$coefficients[5],
      ),
    color = colors[2],
    linewidth = 1.25) +
  # Regression line for Gentoo penguins
  geom_abline(
    aes(
      intercept = lm_combined$fit$coefficients[1] + lm_combined$fit$coefficients[3],
      slope = lm_combined$fit$coefficients[6],
      ),
    color = colors[3],
    linewidth = 1.25)

```

Which, if we now compare to the ggplot drawing, you can see that the biggest
differences are the cut-off in the per-species lines, and the lack of a dark
blue line for the `lm_simple` model.

```{r}

penguins |>
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  ) |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point() +
  geom_smooth(method = lm)

```
